---
title: "Comprehensive multivariate risk modeling improves mortality risk prediction in pulmonary arterial hypertension"
subtitle: "Figures and Tables"
author: "Innsbruck PAH registry"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  pdf_document:
    latex_engine: xelatex
bibliography: pah_biblio.bib
csl: frontiers-in-immunology.csl
header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

library(bookdown)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, dev = "cairo_pdf")


```

\newpage

# Figures {#sec:figures}

```{r fig-consort, fig.width = mm_inch(paper_figures$strobe$w), fig.height = mm_inch(paper_figures$strobe$h), dpi = 600, fig.cap = 'CONSORT flow diagram of the study ana analysis inclusion process.'}

paper_figures$strobe$plot

```

__Figure \@ref(fig:fig-consort). CONSORT flow diagram of the study analysis inclusion process.__

PH: pulmonary hypertension, PHC: right heart catheterization, CTEPH: chronic thromboembolic pulmonary hypertension.

\newpage

```{r fig-uni-cox, fig.width = mm_inch(paper_figures$uni_cox$w), fig.height = mm_inch(paper_figures$uni_cox$h), dpi = 600, fig.cap = 'Factors associated with overall survival identified by univariable Cox modeling.'}

paper_figures$uni_cox$plot

```

__Figure \@ref(fig:fig-uni-cox). Factors associated with overall survival identified by univariable Cox modeling.__

Correlations of `r nrow(pah_study$mod_variables)` candidate variables (__Supplementary Table S1__) with overall survival was investigated with a series of Cox proportional hazard models in the Innsbruck (IBK) and Linz/Vienna (LZ/W) PAH collective (__Supplementary Table S2__). Hazard ratio (HR) significance was assessed with Wald test and corrected for multiple comparisons with Benjamini-Hochberg method. HR with 95% confidence intervals for the factors correlating significantly with survival in at least one cohort are presented in the plot. The range of N number of the complete observations is shown under the plot.

\newpage

```{r fig-sign-os, fig.width = mm_inch(paper_figures$signatures_os$w), fig.height = mm_inch(paper_figures$signatures_os$h), dpi = 600, fig.cap = 'Prediction of overall PH survival by candidate multivariable risk signatures and established risk assessment tools.'}

paper_figures$signatures_os$plot

```

__Figure \@ref(fig:fig-sign-os). Prediction of overall PAH survival by candidate multivariable risk signatures and established risk assessment tools.__

Correlation of the candidate 2 - 4 parameter risk signatures (all possible combinations of `r nrow(pah_study$mod_variables)` variables, __Supplementary Table S1__) with overall survival (OS) in the Innsbruck PAH training cohort (IBK) was investigated by Cox proportional hazard modeling and verified by 20-fold cross-validation. The association of the linear predictor scores for the significant risk signatures was subsequently tested in the Linz/Vienna (LZ/W) cohort by Cox proportional hazard modeling. The significance of model estimates was determined by Wald test, model relevance was assessed by the likelihood ratio test (LRT) and concordance index (C-index). P values were corrected for multiple comparisons with Benjamini-Hochberg method.

__(A)__ Scheme of selection of the developed significant risk signatures.

__(B)__ C-index values with 95% confidence intervals for Cox models of the `r length(multi_modeling$cv_pass$models)` newly developed significant signatures and the established PH risk assessment tools in the training and test cohorts.

\newpage

```{r fig-sign-five, fig.width = mm_inch(paper_figures$signatures_five$w), fig.height = mm_inch(paper_figures$signatures_five$h), dpi = 600, fig.cap = 'Correlation of the newly developed risk signatures with 5-year mortality.'}

paper_figures$signatures_five$plot

```

__Figure \@ref(fig:fig-sign-five). Correlation of the newly developed risk signatures with 5-year mortality.__

The correlation of normalized linear predictor scores (__Supplementary Table S4__) of the `r length(multi_modeling$cv_pass$models)` developed risk signatures significantly associated with overall survival (__Figure \@ref(fig:fig-sign-os)__) with 5-year mortality in the Innsbruck training (IBK) and Linz/Vienna (LZ/W) validation PAH cohorts was investigated by logistic regression. Odds ratio (OR) significance was determined by Wald test and corrected for multiple comparisons with Benjamini-Hochberg method.

__(A)__ Scheme of signature testing.

__(B)__ OR values with 95% confidence intervals for the `r length(multi_modeling$cv_pass$models)` tested signatures and the established PH risk assessment tools in the training and test cohorts.

\newpage

```{r fig-gender-surv, fig.width = mm_inch(paper_figures$gender_surv$w), fig.height = mm_inch(paper_figures$gender_surv$h), dpi = 600, fig.cap = 'The interplay of gender and age affects PAH survival independently of functional classification, six-minute walking distance and NT-pro-BNP levels.'}

paper_figures$gender_surv$plot

```

__Figure \@ref(fig:fig-gender-surv). The interplay of gender and age affects PAH survival independently of functional classification, six-minute walking distance and NT-pro-BNP levels.__

__(A)__ Differences in PAH survival between the participants stratified by age class and sex (IBK: Innsbruck, LZ/W: Linz/Vienna cohort) were assessed by Kaplan-Meier analysis and Mentel-Henszel test. P values corrected for multiple comparisons with Benjamini-Hochberg method are shown in the plots. The numbers of males and females in the age strata are indicated below the plots.

__(B)__  Differences in WHO functional classification (WHO FC), six-minute walking distance (SMWD) and circulating NT-pro-BNP levels at PH diagnosis in the participants stratified by age class and sex were assessed by Kruskal-Wallis test. P values corrected for multiple comparisons with Benjamini-Hochberg method are shown in the plot captions.

\newpage

```{r fig-gender-signatures, fig.width = mm_inch(paper_figures$gender_signatures$w), fig.height = mm_inch(paper_figures$gender_signatures$h), dpi = 600, fig.cap = 'Values of the best performing machine-learning risk signature score in gender and age strata.'}

paper_figures$gender_signatures$plot

```

__Figure \@ref(fig:fig-gender-signatures). Values of the best performing machine-learning risk signature score in gender and age strata.__

The risk signatures predicting overall survival in PH were developed as presented in __Figure 3__. The differences in the signature 2525 linear predictor scores (__Supplementary Table S4__) between the participants stratified by age class and sex (IBK: Innsbruck, LZ/W: Linz/Vienna cohort) were assessed by Kruskal-Wallis test. P values corrected for multiple comparisons with Benjamini-Hochberg method are shown in the plot captions. IBK $\leq$ 60: male n = `r gender$n_numbers$IBK_young$n[1]`, female n = `r gender$n_numbers$IBK_young$n[2]`, IBK >60: male n = `r gender$n_numbers$IBK_elderly$n[1]`, female n = `r gender$n_numbers$IBK_elderly$n[2]`, 
LZ/W $\leq$ 60: male n = `r gender$n_numbers$LZ_young$n[1]`, female n = `r gender$n_numbers$LZ_young$n[2]`, LZ/W >60: male n = `r gender$n_numbers$LZ_elderly$n[1]`, female n = `r gender$n_numbers$LZ_elderly$n[2]`.