```{r fig-strobe, fig.width = 7.086614166, fig.height = 5.511811018, fig.cap = 'CONSORT flow diagram of the study analysis inclusion process.'}

paper_figures$strobe$plot

```

__Figure \@ref(fig:fig-strobe). CONSORT flow diagram of the study analysis inclusion process.__



```{r fig-elanet, fig.width = 7.086614166, fig.height = 7.87401574, fig.cap = 'Multi-parameter survival modeling.'}

paper_figures$elanet$plot

```

__Figure \@ref(fig:fig-elanet). Multi-parameter survival modeling.__



```{r fig-elanet-calibration, fig.width = 7.086614166, fig.height = 3.543307083, fig.cap = 'Elastic net model linear prediction score and overall survival.'}

paper_figures$elanet_calibration$plot

```

__Figure \@ref(fig:fig-elanet-calibration). Elastic net model linear prediction score and overall survival.__



```{r fig-part-clust, fig.width = 7.086614166, fig.height = 7.086614166, fig.cap = 'Clustering of the study participants.'}

paper_figures$part_clust$plot

```

__Figure \@ref(fig:fig-part-clust). Clustering of the study participants.__



```{r fig-cluster-risk, fig.width = 7.086614166, fig.height = 8.267716527, fig.cap = 'Risk assessment and survival differences in the participant clusters.'}

paper_figures$cluster_risk$plot

```

__Figure \@ref(fig:fig-cluster-risk). Risk assessment and survival differences in the participant clusters.__



```{r fig-summary, fig.width = 7.086614166, fig.height = 3.93700787, fig.cap = 'Summary of the analysis results.'}

paper_figures$summary$plot

```

__Figure \@ref(fig:fig-summary). Summary of the analysis results.__

