---
title: "PAH biomarker study"
subtitle: "Figures and Tables"
author: "Innsbruck PAH registry"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  pdf_document:
    latex_engine: xelatex
bibliography: pah_biblio.bib
csl: frontiers-in-immunology.csl
header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

library(bookdown)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, dev = "cairo_pdf")


```

\newpage

# Figures {#sec:figures}

```{r fig-consort, fig.width = mm_inch(paper_figures$uni_cox$w), fig.height = mm_inch(paper_figures$uni_cox$h), dpi = 600, fig.cap = 'CONSORT flow diagram of the study ana analysis inclusion process (placeholder).'}

ggdraw()

```

__Figure \@ref(fig:fig-consort). CONSORT flow diagram of the study ana analysis inclusion process (placeholder).__

\newpage

```{r fig-uni-cox, fig.width = mm_inch(paper_figures$uni_cox$w), fig.height = mm_inch(paper_figures$uni_cox$h), dpi = 600, fig.cap = 'Factors associated with overall survvial identified by univariable Cox modeling.'}

paper_figures$uni_cox$plot

```

__Figure \@ref(fig:fig-uni-cox). Factors associated with overall survvial identified by univariable Cox modeling.__

Correlations of `r nrow(pah_study$mod_variables)` candidate variables (__Supplementary Table S1__) with overall survival was investigated with a series of Cox proportional hazard models in the Innsbruck (IBK) and Linz/Vienna (LZ/W) collective (__Supplementary Table S2__). Hazard ratio (HR) significance was assessed with Wald test and corrected for multiple comparisons with Benjamini-Hochberg method. HR with 95% confidence intervals for the factors correlating significantly with survival in at least one cohort are presented inn the plot. Range of N number of complete observation is shown under the plot.

\newpage

```{r fig-sign-os, fig.width = mm_inch(paper_figures$signatures_os$w), fig.height = mm_inch(paper_figures$signatures_os$h), dpi = 600, fig.cap = 'Prediction of overall PH survival by candidate multivariable risk signatures and established risk assessment tools.'}

paper_figures$signatures_os$plot

```

__Figure \@ref(fig:fig-sign-os). Prediction of overall PH survival by candidate multivariable risk signatures and established risk assessment tools.__

Correlation of the candidate 2 - 4 parameter risk signatures (all possible combinations of `r nrow(pah_study$mod_variables)` variables, __Supplementary Table S1__) with overall survival (OS) in the Innsbruck training cohort (IBK) was investigated by Cox proportional hazard modeling and verified by 20-fold cross-validation. The association of the linear predictor scores for the significant risk signatures was subsequently tested in the Linz/Vienna (LZ/W) cohort by Cox proportional hazard modeling. Significance of model estimates was determined by Wald test, model relevance was assessed by likelihood ratio test (LRT) and concordance index (C-index). P values were corrected for multiple comparisons with Benjamini-Hochberg method.

__(A)__ Scheme of selection of the developed significant risk signatures.

__(B)__ C-index values with 95% confidence intervals for Cox models of the `r length(multi_modeling$cv_pass$models)` developed significant signatures and the established PH risk assessment tools in the training and test cohorts.

\newpage

```{r fig-sign-five, fig.width = mm_inch(paper_figures$signatures_five$w), fig.height = mm_inch(paper_figures$signatures_five$h), dpi = 600, fig.cap = 'Correlation of the developed candidate risk signatures with 5-year mortality.'}

paper_figures$signatures_five$plot

```

__Figure \@ref(fig:fig-sign-five). Correlation of the developed candidate risk signatures with 5-year mortality.__

Correlation of linear predictor scores (__Supplementary Table S5__) of the `r nrow(pah_study$mod_variables)` developed risk signatures significantly associated with overall survival (__Figure \@ref(fig:fig-sign-os)__) with 5-year mortality in the Innsbruck training (IBK) and Linz/Vinna (LZ/W) cohort was investigated by logistic regression. Odds ratio (OR) significance was determined by Wald test and corrected for multiple comparisons with Benjamini-Hochberg method.

__(A)__ Scheme of signature testing.

__(B)__ OR values with 95% confidence intervals or Cox models of the `r length(multi_modeling$cv_pass$models)` tested signatures and the established PH risk assessment tools in the training and test cohorts.

\newpage

```{r fig-clustering, fig.width = mm_inch(paper_figures$clustering$w), fig.height = mm_inch(paper_figures$clustering$h), dpi = 600, fig.cap = 'Distinct risk phenotypes of PH defined by acute mortality, motility, daily functioning and age and sex.'}

paper_figures$clustering$plot

```

__Figure \@ref(fig:fig-clustering). Distinct risk phenotypes of PH defined by acute mortality, motility, daily functioning and age and sex.__

PH patients in the Innsbruck (IBK) and Linz/Vienna cohort (LZ/W) were subjected to association analysis in respect to age at diagnosis, sex (0: male), six-minute walking distance (SMWD), WHO functional classification and 5-year mortality (0: survivor) using DBSCAN algorithm. Three subsets were identified (Subset 1 - 3, UA: unassigned).

__(A)__ Subset assignment displayed in a score plot of results of two-dimensional PCA (principal component analysis) in respect to the clustering variables. PC: principal component, percents of variation associated with the given PC is presented in the axes. Cluster n numbers are presented below the plots.

__(B)__ Comparison of minimum/maximum normalized levels of the clustering variables between the subsets.

__(C)__ Survival differences between the subsets investigated by Kaplan-Meier analysis and log-rank test. OS: overall survival. Test p values are shown in the plots.

\newpage

```{r fig-cluster-score, fig.width = mm_inch(paper_figures$cluster_risk_scales$w), fig.height = mm_inch(paper_figures$cluster_risk_scales$h), dpi = 600, fig.cap = 'Newly developed risk signatures differentiate between age, sex, motility, daily functioning and mortality-defined risk phenotypes.'}

paper_figures$cluster_risk_scales$plot

```

__Figure \@ref(fig:fig-cluster-score). Newly developed risk signatures differentiate between age, sex, motility, daily functioning and mortality-defined risk phenotypes.__

Linear predictor score values (__Supplementary Table S5__) for the best performing risk signatures 309 (age, cardiac index and sex) and 2525 (age, cardiac index, mPAP and sex) and the established PH risk assessment tools were compared between the PH risk subsets defined in __Figure \@ref(fig:fig-clustering)__ with Kruskal-Wallis test. P values were corrected for multiple comparisons with Benjamini-Hochberg method.

__(A)__ Normalized score medians in the risk subsets presented as a heat map.

__(B)__ Kruskal-Wallis test p values. Orange: significant comparisons.